---
title: "2022 Formula One Drivers' World Championship"
author: '22723583 Liu Shuai'
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load packages required for all of your analysis here
library(readr)
library(tidyverse)
library(gganimate)
library(gifski)
library(ggrepel)
```

Change the title and name above. Use your student number as your name.

If you are using shiny or gganimate change output to HTML

## Data Background 

Write 1 or 2 paragraphs on what is interesting about your data

Include reference [link](http://www.link.com) to the location of the data set that you are going to clean and visualize.

### Content
The dataset consists of all information on the Formula 1 races, drivers, constructors, qualifying, circuits, lap times, pit stops, championships from 1950 till the latest 2022 season.

### Link
[Formula 1 Race Data](https://www.kaggle.com/datasets/jtrotman/formula-1-race-data)


## Data Tidying Code

```{r}
# data import code
races <- read_csv("./data-raw/races.csv")
results <- read_csv("./data-raw/results.csv")
drivers <- read_csv("./data-raw/drivers.csv")
sprint <- read_csv("./data-raw/sprint_results.csv")
# data tidying code (ready for plot later)
# filter F1 2022 season 
str(races$year)
races <- races %>% 
  filter(year == 2022) %>% 
  select(raceId:date,-c("year","circuitId"))
head(races)

# combine results with drivers,add drivers name
drivers <- drivers %>% 
  # choose coloum raceId,Driver's forename and surname,code
  select(1,4:6) 

# rename name column in races data
races <- races %>% 
  rename(Grand_Prix = name)

# replace string "Grand Prix" in column Grand_Prix with "GP"
races$Grand_Prix <- races$Grand_Prix %>% 
  str_replace(pattern = "Grand Prix",replacement = "GP")
races$Grand_Prix


# choose F1 2022 season race results
results <- results %>% 
  filter(raceId >= 1074) %>% 
  select(1:3,10,18)

# combine races results with races
races <- left_join(races,results,by = "raceId")
str(races)

# combine races with drivers
races <- left_join(races,drivers,by = "driverId")

# combine sprint races results with race results 
glimpse(sprint)
glimpse(results)

sprint <- sprint %>% 
  # choose 2022 season sprint races
  filter(raceId >= 1074) %>% 
  # select "points","raceId" and "driverId" column,rename "points" column to "sprintpoint"
  transmute(sprintpoint = points,raceId,driverId) 
# join races date with sprint races date using left_join
races <- left_join(races,sprint,by = c("raceId","driverId"))

# clean sprint race points' missing data
sum(is.na(races$sprintpoint))
# replace missing values with 0
races$sprintpoint <- replace_na(races$sprintpoint,
                                 replace = 0)
# check if there have missing values
sum(is.na(races$sprintpoint))

# calculate each driver's total points (= sprint points + races points)
races <- races %>% 
  group_by(surname) %>% 
  arrange(round) %>% 
  mutate(totalpoint = sprintpoint + points)

# calculate each driver's cumulative points
races <- races %>% 
  group_by(surname) %>% 
  arrange(round) %>% 
  # calculate cumulative points and creat new column named cumpoint 
  mutate(cumpoint = if_else(is.na(lag(totalpoint)),true = totalpoint,false = cumsum(totalpoint)))

# check cumulative points
races %>% 
  group_by(surname,round) %>% 
  summarise(totalpoint,cumpoint)
# use lots of comments to help us show that you understand what you doing with your code... which will get you a higher mark.
```

## Visualisation Code and Plot

Write a paragraph about your plot

As 2022 season Formula One drew to a close after Abu Dhabi Grand Prix,it's time to review the impressive season.

1. Ferrari and Leclerc behaved well at the start of this season, while the final drivers' world championship belongs to Verstappen. How did that happened? What was the turning point of it?

2. Four-time world champion Sebastian Vettel announced his retirement. How he performed in his last season?

3. Zhou Guanyu is the first Chinese driver who currently competes in Formula One for Alfa Romeo. What is performance in the rookie year? How his teammate Bottas behaved in the season?

To answer these questions,I will draw an animate trend picture fo 2022 season Formula One Drivers' World Championship.

## Results
1. The driver's championship turning point was Spanish Grand Prix. Verstappen had the edge over Leclerc and other drivers after that. It was a cut throat competition for P2.

2. Vettel's performance was good for his last season.

3. Both Zhou and Bottas had a good start in the opening game, but it was only an average performance at the end of season. Though they earned less points in the second half period, the final results is a huge improvement for Alpha Romeo.
```{r}
# plot code
# choose specific drivers 
# Verstappen(VER),Pérez(PER),Leclerc(LEC),Norris(NOR),Vettel(VET),Bottas(BOT),Zhou(ZHO)
driverlab <- races %>% 
  filter(code %in% c("VER","LEC","PER",
                     "NOR","BOT","VET","ZHO"))

# specify hex color for drivers (labels)
# Red Bull racing blue for Verstappen and Pérez
# Ferrari red for Leclerc
# McLaren orange for Norris
# Aston Martin dark green for Vettel
# Alfa Romeo red for Bottas
# Alfa Romeo green for Zhou
cols <- c("VER" = "#23326A","LEC" = "#CD212A",
          "PER" = "#23326A","NOR" = "#FF8000",
          "BOT" = "#A42134","VET" = "#002420",
          "ZHO" = "#1E5341")

g <- ggplot(data = races,mapping = aes(x = round,
                                       y = cumpoint)) +
  # plot line plots
  geom_line(aes(colour = code,alpha = 0.5)) + 
  # add scatter plots to line plots
  geom_point(aes(colour = code,alpha = 0.5)) +
  # add vertical line
  geom_vline(xintercept = c(5.5,12.5),color = "Gray50",alpha = 0.5,linetype = 6) +
  # Scale
  # set chosen color for specific drivers,gary color for other drivers 
  scale_colour_manual(values = cols,na.value = "Gray50") +
  # substitute x axis(round) with Grand Prix's name (Grand_Prix)
  scale_x_continuous(breaks = races$round,
                     labels = races$Grand_Prix,
                     # put start point at (0,0)
                     expand = c(0,0)) +
  # plot y on square root scale to avoid dense line plot,because most driver's points are between 0 and 200,while some are above 400
  scale_y_sqrt() +
  # let labels shows outside the coordinate
  coord_cartesian(clip = 'off') + 
  # Labels
  # add x,y axis/caption title
  labs(x = "Grand Prix",
       y = "Points",
       # use \n for a newline
       caption = "Source:Formula 1 Race Data \n https://www.kaggle.com/datasets/jtrotman/formula-1-race-data",
       title = "2022 Formula One Drivers' World Championship") +
  # Theme
  theme_classic() +
  theme(plot.title = element_text(face = "bold.italic"),
        plot.caption = element_text(face = "italic"),
        # rotate x axis labels to avoid overlapped
        axis.text.x = element_text(angle = 45, hjust = 1),
        # hide legend
        legend.position = "none") +
  # annotate
  # add labels for vertical lines
  annotate("text",x = c(5.2,12.5),y = c(300,425),
           label = c("drivers' \n championship \n turning \n point",
                "Vettel's retire \n announcement"),
           size = 3,
           fontface = "italic")
g

h <- g + 
  # add labels for chosen drivers,use ggrepel package to avoid overlapped labels
  geom_label_repel(aes(x = driverlab$round,
                       group = driverlab$code,
                       # set dashed line  
                       segment.linetype = 2,
                       # set color for each labels
                       colour = driverlab$code),
                       label = driverlab$code,
                       # set labels for chosen drivers
                       data = subset(driverlab),
                       # expand the scale to make room for labels
                       nudge_x = 3,
                       # adjust labels in y (axis) direction
                       direction = "y",
                       # set seeds to avoid random adjustment
                       seed = 100,
                       hjust = 0) +
  # choose color for data labels manually
  scale_colour_manual(values = cols) +
  transition_reveal(round) 

animate(h,
        # set gif duration to 30s
        duration = 30,
        height = 500, 
        width =800)

anim_save("2022 Formula One Drivers' World Championship.gif")
# again, make sure you include lots of comments to let us know that you know what the code is doing.
```
